---
title: "Work Sample"
author: "Micah James"
date: "December 20, 2017"
output: html_document
---

# The Problem

Suppose that you were given a spreadsheet that contained the birth year and year
of death for a list of people. Suppose also that we wish to determine the year
in which the greatest number of these people had lived. In this post, I'll show a couple different ways one could find the answer to this question using R.

# Some Preliminaries

To begin, we'll load some packages. If you don't already have these installed, I
highly recommend you install the meta package `tidyverse` to get `dplyr`,
`stringr`, `tidyr` and many other useful packages. We also load `microbenchmark` so that we can have a discussion about runtimes for the our two solutions.

```{r "Package Loading"}
# install.packages("tidyverse")
library(dplyr)
library(stringr)
library(tidyr)
#install.packages("microbenchmark")
library(microbenchmark)
```

I don't have a spreadsheet like the one mentioned in the problem so we'll
generate one here. It's not particularly exciting but it does its job and
produces a csv with nearly 20,000 rows in it. To avoid recreating that file
every time this code is run, it first checks to make sure we haven't already
produced `lifetimes.csv` in the current working directory.

```{r "Creating The Data Set"} 
if (!file.exists("lifetimes.csv")){
  set.seed(11111)
  samplesize <- 200000
  data_frame(
    id = 1:samplesize,
    birthyear = 
      sample(
        x = 1901:1999,
        size = length(id),
        replace = TRUE
      ),
    endyear = birthyear + rgeom(n = length(id),prob = 0.002)
    ) %>% 
    filter(endyear < 2000) %>% 
    write.csv("lifetimes.csv",row.names=FALSE)
}
```

```{r "import data set", echo=FALSE}
lifetimes <- 
  read.csv(
    "lifetimes.csv",
    stringsAsFactors = FALSE
    ) %>% 
  as_tibble()
```

Here's what `lifetimes` looks like. Note that I converted it to a tibble for
nicer printing.

```{r "lifetimes"}
head(lifetimes) # Imported from "lifetimes.csv"
```

Our first solution relies on R's `table` function to do the heavy (and fast)
lifting. The `table` function applied to a vector will give a frequency table of
the elements of that vector. The table is returned as a named vector with the
names equal to the element that is counted and the vector entry is the number of
times that element appeared.

```{r "Table Demonstrated"}
Group1 <- c("Alice","Bob","Bob","Bob","Jim","George","George")
Group2 <- c("Alice","Bob","Jim","Jim","George")
table(Group1)
# Note how the names are sorted in the output.
table(Group2)
table(Group1) + table(Group2)
```
Note that R makes no effort to match up names when adding two tables together.
What's more, when adding two tables, the names of the first table in the sum becomes the names of the sum.
```{r "More table demonstration"}
Group3 <- c("George","Hadley","Leon","Mara")
table(Group3)
table(Group2) + table(Group3)
```

So while the `table` function will do compute our frequencies quickly, we're
going to rely on dataframes to ensure that the frequencies keep their
association with their names.

# Our Clever Solution

In our first solution, we'll convert the output of the table function into
dataframes to preserve the frequencies' associations with their years. The
clever bit here, is that instead of counting the frequencies of the end years,
we're going to count the frequencies of the end years plus 1. Consider the
person in the list with id `r lifetimes$id[1]`. That person has a listed endyear
of `r lifetimes$endyear[1]` so they presumably were living at the beginning of
`r lifetimes$endyear[1]`. It is in year `r lifetimes$endyear[1]+1` when they
stop being counted among the living.

For each year from 1901 through 2000, we will compute the net change in the
number of people living by adding the number of people born that year and
subtracting the number of people who died in the previous year. We then will
compute the cumulative sum of the net changes to compute the number of people
from this list who were alive at some point in the given year.

```{r "First Solution"}
addcounts <- function(df){
  sum(df$counts)
}

ComputeNetChanges <- function(firstyears, endyears){
  firstyearcounts <- 
    table(firstyears)
  FYDataFrame <- 
    data_frame(
      year = str_c("Y", names(firstyearcounts), sep=""),
      label = "first",
      counts = firstyearcounts
    )
  yearafterendyearcounts <- table(endyears+1)
  EYDataFrame <- 
    data_frame(
      year = str_c("Y", names(yearafterendyearcounts), sep = ""),
      label = "end",
      counts = -yearafterendyearcounts
    )
  CombinedDataFrame <- 
    bind_rows(FYDataFrame,EYDataFrame) %>% 
    group_by(year) %>%
    nest() %>%
    mutate(
      netchange =
        purrr::map(
          data,
          addcounts
        ) %>% unlist(),
      numberlived = cumsum(netchange)
    ) %>%
    arrange(desc(numberlived))
  CombinedDataFrame
}
microbenchmark(
  ComputeNetChanges(lifetimes$birthyear, lifetimes$endyear),
  times=10
)
finalresults <- ComputeNetChanges(lifetimes$birthyear, lifetimes$endyear)
head(finalresults)
```

As we can see, the year in which the greatest number of these people lived was
in `r finalresults$year[1]` with `r finalresults$numberlived[1]` different
people from the list alive at some point during that year.

I consider this solution to be the more clever of the two solutions because,
apart from the work that the `table` function does for us, each of these other
operations only involves a single pass through the tibble containing the birth
and end years for the people in our dataset. This should save on time and memory
usage.

# A Simpler, Slower Solution

A simpler solution would have been to replace each person with a list of years
in which they lived. Combining these lists creates a vector in which each number
represents a single person living in that year. The question then becomes: which
year occurs most frequently in this list of years?

```{r "brute force solution"}
YearsAlive <- function(firstyears,endyears){
  purrr::map2(
    .x = firstyears,
    .y = endyears,
    seq
  ) %>% unlist %>% table %>% sort(decreasing=TRUE)
}

microbenchmark(
  YearsAlive(lifetimes$birthyear,lifetimes$endyear),
  times=10
)
YearsAlive(lifetimes$birthyear,lifetimes$endyear)[1:6]
```

Note that we relied on the `table` function again this time, but only once at
the end and only with its intended use. This solution takes, on average, 3 times
longer on my machine.
